---
title: "Results"
author: "Matthew Wankiewicz"
date: '2022-12-14'
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(sjPlot)
deploy <- read_csv("deploy.csv")
model <- read_rds("final_model.rds")
```

## Question 1

```{r model prediction}
deploy$estimate <- predict(model, newdata = deploy,
                           type = "response")
write_csv(deploy, "deploy.csv")
```

## Question 2

After loading in the data, I found that there were some missing values in the `SpinRate` column. Instead of omitting them, I decided to replace them with the average spin rates in the data. After doing this, I split the data into a training and testing set and started to create models. The two types of models I focused on were logistic regression and Random Forest, since we are predicting a binary response. After trying many different variable combinations, I settled on a logistic regression model because it was able to create estimates that made sense and was accurate in terms of the data, but was also able to quantify the impact of each variable on the chances of the ball being put in play. I also attempted to balance the data as roughly 70% of the `InPlay` column were 0's. After over-sampling the data and re-fitting the models, I found that the results were better with the model built prior to over-sampling. When testing the models, I used confusion matrices and ROC curves to determine which models captured the given data best.

## Question 3

```{r model display, message=FALSE, echo=FALSE}
plot_model(model) +
  labs(subtitle = "Red variables decrease contact probability, blue variables increase") +
  scale_y_continuous(limits = c(0.5, 2.25))
```

Of the four variables of interest, we see that higher velocities and higher Induced Vertical Breaks lead to an increase in the chance of a ball being put in play. On the other hand, we see that having a higher horizontal break leads to a decreased chance of a ball being put in play, the same goes for pitches with higher spins which decrease the probability but a lesser rate.

## Question 4

I would want to look at the results with left-handed hitters included in the data as well, it's highly likely that the pitcher will end up facing a lefty and we would need to see if these results stay true or change. I would also like to see what the situation was when each pitch was being thrown, if the count was 3-0, the batter will not want to swing, but in 0-2 or 1-2, they'll have to go for anything close.

## Question 5

All coding is contained the `models.R` file. This contains data visualization, data cleaning and model creation/testing.
